Dynamical modeling of tidal streams

### 2013-09-20 tex/ms.tex
streams: further work on action-angle framework

\documentclass[12pt,preprint]{aastex}
%\documentclass{emulateapj}
\usepackage{amssymb,amsmath}
%\usepackage{caption}
%\DeclareCaptionLabelSeparator{dotemdash}{.--- }
%\captionsetup[figure]{labelformat=simple, labelsep=dotemdash}
%\usepackage{color,hyperref}
% hypertex insanity
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\newcounter{address}
\setcounter{address}{1}
%\usepackage{sidecap}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\equationname}{\eqnname}
\renewcommand{\tablename}{Table}
\renewcommand{\figurename}{Figure}
\newcommand{\figurenames}{\figurename~s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\normal}{\ensuremath{\mathcal{N}}}
\newcommand{\flag}[1]{\texttt{\lowercase{#1}}}
\newcommand{\feh}{\ensuremath{[\mathrm{Fe/H}]}}
\newcommand{\afe}{\ensuremath{[\alpha\mathrm{/Fe}]}}
\newcommand{\logg}{log g}
\newcommand{\Ro}{\ensuremath{R_0}}

\renewcommand{\vec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\vecx}{\ensuremath{\vec{x}}}
\newcommand{\vecv}{\ensuremath{\vec{v}}}
\newcommand{\vech}{\ensuremath{\vec{H}}}
\newcommand{\vecj}{\ensuremath{\vec{J}}}
\newcommand{\veco}{\ensuremath{\vec{\Omega}}}
\newcommand{\veca}{\ensuremath{\boldsymbol\theta}}
\newcommand{\df}{\ensuremath{f}}
\newcommand{\tdf}{\ensuremath{\mathrm{tDF}}}
\newcommand{\paramsdiff}{\ensuremath{\vec{p}_\Delta}}
\newcommand{\paramsdf}{\ensuremath{\vec{p}_{\mathrm{DF}}}}
\newcommand{\paramspot}{\ensuremath{\vec{p}_\Phi}}

@@

@@

The probability in \eqnname~(\ref{eq:aadfsimple}) incorporates and
elucidates commonly used methods for fitting tidal streams. On the one
hand are approaches that search to minimize the spread in energy,
integrals of the motion, or actions (\eg,
\citealt{Binney08a,Penarrubia12a}, Sanderson \etal, 2013, in
preparation). These approaches ignore the correlations between actions
and angles in the stream (the exponential in
\eqnname~(\ref{eq:aadfsimple}), which describes this correlation
through the frequencies) and only use $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$; finding the gravitational potential that
minimizes the spread in actions optimizes $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$. Recently, a method was proposed that only uses
the correlation between the actions and the angles, expressed by the
exponential in \eqnname~(\ref{eq:aadfsimple}) \citep{Sanders12a}; this
method does not use the fact that the spread in actions in a tidal
stream is small (expressed by $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$).

>>>>

\documentclass[12pt,preprint]{aastex}
%\documentclass{emulateapj}
\usepackage{amssymb,amsmath}
%\usepackage{caption}
%\DeclareCaptionLabelSeparator{dotemdash}{.--- }
%\captionsetup[figure]{labelformat=simple, labelsep=dotemdash}
%\usepackage{color,hyperref}
% hypertex insanity
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\newcounter{address}
\setcounter{address}{1}
%\usepackage{sidecap}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\Eqnname}{Equation}
\newcommand{\equationname}{\eqnname}
\renewcommand{\tablename}{Table}
\renewcommand{\figurename}{Figure}
\newcommand{\figurenames}{\figurename~s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\normal}{\ensuremath{\mathcal{N}}}
\newcommand{\flag}[1]{\texttt{\lowercase{#1}}}
\newcommand{\feh}{\ensuremath{[\mathrm{Fe/H}]}}
\newcommand{\afe}{\ensuremath{[\alpha\mathrm{/Fe}]}}
\newcommand{\logg}{log g}
\newcommand{\Ro}{\ensuremath{R_0}}

\renewcommand{\vec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\unitvec}[1]{\ensuremath{\mathbf{\hat{#1}}}}
\newcommand{\vecx}{\ensuremath{\vec{x}}}
\newcommand{\vecv}{\ensuremath{\vec{v}}}
\newcommand{\vech}{\ensuremath{\vec{H}}}
\newcommand{\vecj}{\ensuremath{\vec{J}}}
\newcommand{\veco}{\ensuremath{\vec{\Omega}}}
\newcommand{\veca}{\ensuremath{\boldsymbol\theta}}
\newcommand{\df}{\ensuremath{f}}
\newcommand{\tdf}{\ensuremath{\mathrm{tDF}}}
\newcommand{\paramsdiff}{\ensuremath{\vec{p}_\Delta}}
\newcommand{\paramsdf}{\ensuremath{\vec{p}_{\mathrm{DF}}}}
\newcommand{\paramspot}{\ensuremath{\vec{p}_\Phi}}

@@

The discussion so far has been entirely general and applies to any
sort of stellar debris that originates from a single orbit, without
necessarily generating a one-dimensional structure. In most
observations of tidal streams, however, they appear as one-dimensional
objects rather than the six-dimensional objects that we have
considered so far. The reason for this is both physical and
observational: (a) narrow tidal streams form because a star separates
from the progenitor over time mostly along a single direction in
angle-space: the Hessian matrix of the Hamiltonian, $\partial^2 H /
\partial \vecj \partial \vecj$, has a single large eigenvalue, which
determines the direction along which the stream is disrupted in
angle-space (because $\partial \veco / \partial \vecj$ is equal to the
Hessian); (b) because many tidal streams are dynamically cold with
velocity dispersions $\lesssim10\kms$, observations are unable to
resolve the angle differences along the two eigenvectors of the
Hessian corresponding to the two smallest eigenvalues or the spread in
actions among stream members. If the latter can be resolved, the
stream is four-dimensional. In what follows we will be mainly
concerned with the stream only spreading along a single direction in
angle space.

If the stream is one-dimensional in angle space then the time since
the removal from the progenitor becomes unobservable if the progenitor
orbit is unknown: above, determining the time depended on explaining
the three-dimensional angle difference $\Delta \veca$ with the
three-dimensional frequency difference $\Delta \veco$ using a single
time difference; if two of the angle and frequency differences are
unable to be resolved, then the frequency difference and the time
difference become degenerate (aka, you can draw a straight line
connecting any two points). To make use of this fact to simplify the
dynamical model, we write the frequency difference as
\begin{equation}
  \Delta \veco = \Delta \Omega_1\,\unitvec{e}_1\,,
\end{equation}
where $\unitvec{e}_1$ is a unit vector. Then we write the angle
difference $\Delta \veca$ 
\begin{equation}
  \Delta \veca = \Delta \theta_1\,\unitvec{e}_1 + \Delta \veca_\perp\,\unitvec{e}_1^\perp\,,
\end{equation}
where$\unitvec{e}_1^\perp$ is the orthogonal complement of
$\unitvec{e}_1$ (the two directions perpendicular to $\unitvec{e}_1$)
and $\Delta \veca_\perp$ are the components of $\Delta \veca$ in that
direction. We can then write \eqnname~(\ref{eq:aadfsimple}) as
\begin{equation}\label{eq:aadfsimpler}
\begin{split}
  p(\vecx_i,\vecv_i & | \Phi,\vecj_s,\veca_s,\paramsdiff) 
   = \\
   & f_{\vecj}(\Delta \vecj_i|\paramsdiff) \,  \frac{1}{2\,\pi\,\sigma_\theta^2\,\sqrt{\sum_{R,\phi,Z}\Delta \veco_i^2}}\,\exp\left(-\frac{|\Delta\veca_{i,\perp}|^2}{2\,\sigma_\theta^2}\right)\,.
\end{split}
\end{equation}
\Eqnname~(\ref{eq:aadfsimpler}) is equivalent to
\eqnname~(\ref{eq:aadfsimple}), but it expresses the dynamical model
in a slightly different way. Whereas the angle part of
\eqnname~(\ref{eq:aadfsimple}) allows the correct potential and
progenitor to be inferred by asking whether a single time difference
explains the three angle differences given the three frequencies,
\eqnname~(\ref{eq:aadfsimpler}) `rewinds' the star's orbit along the
direction of $\Delta \veco$ and asks whether the angle difference
perpendicular to $\Delta \veco$ is consistent with the initial angle
spread (because it has not evolved since the removal of the star). In
re-writing this, we have not yet used the approximation that the
stream only forms along a single direction in $\Delta \veco$.

@@

The probability in \eqnname~(\ref{eq:aadfsimple}) incorporates and
elucidates commonly used methods for fitting tidal streams. On the one
hand are approaches that search to minimize the spread in energy,
integrals of the motion, or actions (\eg,
\citealt{Binney08a,Penarrubia12a}, Sanderson \etal, 2013, in
preparation). These approaches ignore the correlations between actions
and angles in the stream (the exponential in
\eqnname~(\ref{eq:aadfsimple}), which describes this correlation
through the frequencies) and only use $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$; finding the gravitational potential that
minimizes the spread in actions optimizes $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$. Recently, a method was proposed that only uses
the correlation between the actions and the angles, expressed by the
exponential in \eqnname~(\ref{eq:aadfsimple}) \citep{Sanders13a}; this
method does not use the fact that the spread in actions in a tidal
stream is small (expressed by $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$).
